---
title: "Reproducible Research: Peer Assessment 1"
author: Jenny Eriksson
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
This section will load the data and transform the date feature to a date format.
```{r}
unzip("activity.zip")
data<-read.csv('activity.csv')
data<-transform(data, date=as.Date(date))
```


## What is mean total number of steps taken per day?
By aggregating the data by day we can get the total number of steps as shown in the table below.

```{r, results="asis"}
nSteps<-aggregate(steps~date, data=data, sum)
library(xtable)
xt<-xtable(transform(nSteps, date=as.character(date)))
print(xt, type="html")
```
We can also make a histogram showing the freuquency of steps per day, see below.
```{r}
hist(nSteps$steps, main="Histogram: number of steps taken per day", xlab="number of steps [count]", ylab="frequency [days]", ylim=c(0,35))

```

The mean and median is computed as
```{r}
meanSteps <- format(round(mean(nSteps$steps),1), nsmall=1)
medianSteps <- format(round(median(nSteps$steps),1),nsmall=1)

```
where we get a mean of `r meanSteps` steps/day and a median of `r medianSteps` steps/day.

## What is the average daily activity pattern?
By instead aggregating on the interval we get the mean number of steps per interval, see the plot below.
```{r}
meanStepsInterval<-aggregate(steps~interval, data=data, mean)
plot(meanStepsInterval$interval, meanStepsInterval$steps, type="l", main="Average number of steps per interval", xlab="interval", ylab="steps [count]") 
```
The maximum number of steps taken, as a mean, is
```{r}
idx<-which.max(meanStepsInterval$steps)
maxStep<-format(round(meanStepsInterval$steps[idx],1),nsmall=1) 
maxStepInterval<-meanStepsInterval$interval[idx]
```
`r maxStep` steps found at interval `r maxStepInterval`, which corresponds to the time 8.35AM.


## Imputing missing values
In this data set there are some missing values, found as
```{r}
sum(is.na(data$steps))
```

In order to get a more accurate analysis of the data we will impute the missing values with the mean value of the corresponding interval, which has been computed earlier in `meanSteps`. The code snippet below will loop through all rows of `data` and create a new table, `dataImputed`, that will contain imputed values.
```{r}
dataImputed <- data
for(i in 1:nrow(dataImputed))
{
    if(is.na(dataImputed$steps[i]))
    {
        replacementvalue <- meanStepsInterval[meanStepsInterval$interval==dataImputed$interval[i],2]
        dataImputed$steps[i]<-replacementvalue
    }
}

```
The histogram of the new table is computed and shown belown. We see a higher frequency in the range 10 000-15 000 steps per day, which makes sense. Since this category is the most common one, it's probable that the replacement values for each interval will contribute most to this category.
```{r}
nStepsImputed<-aggregate(steps~date, data=dataImputed, sum)
hist(nStepsImputed$steps, main="Histogram: number of steps taken per day, Imputed data", xlab="number of steps [count]", ylab="frequency [days]")

```

The mean and median is updated
```{r}
meanStepsImputed <- format(round(mean(nStepsImputed$steps),1), nsmall=1)
medianStepsImputed <- format(round(median(nStepsImputed$steps),1),nsmall=1)

```
where we get the same mean as before `r meanStepsImputed`/day,which is explainable by the fact that we are replacing with mean values. The median is slightly higher, `r medianStepsImputed`/day. The explanation is that we replaced with mean values, which were higher than the (previous) median values, hence the median is increased.


## Are there differences in activity patterns between weekdays and weekends?
To answer the question on how the activity changes over the week we will construct a new column, indicating whether it is the weekend or not, see code below.
```{r}
Sys.setlocale(category = "LC_ALL", locale = "english")
dataImputed <- cbind(dataImputed, weekend=factor(weekdays(dataImputed$date) %in% c("Saturday", "Sunday"), labels = c("Weekday", "Weekend")))
head(dataImputed)
``` 

We can now view the changes in activity by aggregating on weekend/weekday
```{r}
library(ggplot2)

meanStepsIntervalImputed<-aggregate(steps~interval+weekend, data=dataImputed, mean)

qplot(interval, steps, data=meanStepsIntervalImputed, col=weekend, geom="line", main="Average number of steps per interval", xlab="interval", ylab="steps [count]") 
```
 When comparing the activity we find that during the weekdays the person tend to walk more in the morning and less in the afternoon, compared to weekends.

